When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-10-16 doc/paper/ms.tex
updated to fix typos in prrof

\section{Methods}
\label{methods}
The Penman-Monteith equation \cite<hereafter PM,>[]{Penman_1948,
  Monteith_1965} estimates ET as a function of observable atmospheric
variables and surface conductances:
% \begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R_{net} + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
% \end{linenomath*}
  where $\Delta$ is the change in saturation vapor pressure with
  temperature, given by Clausius-Clapeyron
  ($\frac{d \; e_s}{d \; T}$), $R_{net}$ is the net radiation minus
  ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air
  density, $c_p$ is specific heat of air at constant pressure,
  $\gamma$ is the psychometric constant, and $g_s$ is the stomatal
  conductance (Table \ref{definitions}). The issue with PM is that
  $g_s$ is a function of carbon uptake, which is has a strong
  functionally relation to ET through stomatal function. Therefore,
  when PM is formulated in terms of $g_s$, it is an implicit function
  of ET itself rather than an explicit function of ET. Here we will
  derive a new form of PM in which ET is an explicit function of plant
  parameters and environmental conditions, and use it to assess the
  ecosystem scale response to VPD (by taking a partial derivative).

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \footnotesize
  \begin{adjustbox}{center=15cm}
    \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citeA{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_{s}$  &  stomatal conductance & m s$^{-1}$
                                   & \citeA{Medlyn_2017} \\
    $g_{1}$  & ecosystem-scale slope parameter & Pa$^{0.5}$ & \citeA{Medlyn_2017} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda = \frac{\partial \; transpiration}{\partial\; CO_2\; assimilation}$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration &
                                                                      - & - \\
    GPP & gross primary production & $\mu$-mol C s$^{-1}$ m$^{-2}$ & -
    \\
    ET & evapotranspiration & W m$^{-2}$ & - \\
   uWUE & underlying water use efficiency & $\mu$-mol C Pa$^{0.5}$
                                            J$^{-1}$ ET  &
                                                           \citeA{Zhou_2015} \\
    \hline
  \end{tabular}
\end{adjustbox}
\end{table}

@@

\begin{table}
  \caption{Variable quantities in the ET response to VPD. Each value
    is varied to determine the effect of a range of expected plant and
    environmental conditions on ET response to VPD. A citation is
    provided for values of g$_1$ and uWUE, which are directly derived
    from previous literature. Conceptually, min. values are extracted
    from literature to correspond to approximately the 15th percentile
    of observed conditions during the growing season, med. values
    correspond to approximately the 50th percentile, and max. values
    correspond to approximately the 85th percentile. Values for $T$
    and g$_a$ are calculated from FLUXNET-2015 data (see supplemental
    material for description), rounding the 15th, 50th, and 85th
    percentile to the nearest 5$^o$C and 0.005 m/s, respectively.}
  \label{param_varying}
  \centering
  \footnotesize
  \begin{adjustbox}{center=15cm}
  \begin{tabular}{l c c c c c}
    \hline
    Symbol & units [units in citation] & min  & med & max
    & citation  \\
    \hline
g$_1$ & Pa$^{1/2}$ [kPa$^{1/2}$] & 63.25 [2.00] & 126.49 [4.00] & 189.74 [6.00] & Fig. 2, 7; \citeA{Medlyn_2017} \\
uWUE & $\mu$-mol C Pa$^{0.5}$ J$^{-1}$ ET [g C hPa$^{1/2}$ kg$^{-1}$ H$_2$O] & 2.33 [6.99] & 3.17 [9.52] & 4.01 [12.05] & Table 4; \citeA{Zhou_2015} \\
T & $^o$C & 10.00 & 20.00 & 30.00 & - \\
g$_a$ & m/s & 0.015 & 0.035 & 0.055 & - \\
    \hline
  \end{tabular}
  \end{adjustbox}
\end{table}

@@

\acknowledgments This work used eddy covariance data acquired and shared
by the FLUXNET community, including these networks: AmeriFlux, AfriFlux,
AsiaFlux, CarboAfrica, CarboEuropeIP, CarboItaly, CarboMont,
ChinaFlux, Fluxnet-Canada, GreenGrass, ICOS, KoFlux, LBA, NECC,
OzFlux-TERN, TCOS-Siberia, and USCCC. The ERA-Interim reanalysis data
are provided by ECMWF and processed by LSCE. The FLUXNET eddy covariance
data processing and harmonization was carried out by the European Fluxes
Database Cluster, AmeriFlux Management Project, and Fluxdata project of
FLUXNET, with the support of CDIAC and ICOS Ecosystem Thematic Center,
and the OzFlux, ChinaFlux and AsiaFlux offices. We would like to thank
Dr. Trevor Keenan for providing tools for citation of the FLUXNET2015
dataset, available at
\url{https://github.com/trevorkeenan/FLUXNET_citations}. This material is
based upon work supported by the National Science Foundation Graduate
Research Fellowship under Grant No. DGE 16-44869. All code and data used
in this analysis, including those used to generate the figures and
tables, are publicly available at
\url{https://github.com/massma/climate\_et}. We would
like to thank Emma Robinson and one anonymous reviewer for thoughtful
comments that greatly improved the strength of the manuscript.

>>>>

\section{Methods}
\label{methods}
The Penman-Monteith equation \cite<hereafter PM,>[]{Penman_1948,
  Monteith_1965} estimates ET as a function of observable atmospheric
variables and surface conductances:
% \begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R_{net} + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
% \end{linenomath*}
  where $\Delta$ is the change in saturation vapor pressure with
  temperature, given by Clausius-Clapeyron
  ($\frac{d \; e_s}{d \; T}$), $R_{net}$ is the net radiation minus
  ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air
  density, $c_p$ is specific heat of air at constant pressure,
  $\gamma$ is the psychometric constant, and $g_s$ is the stomatal
  conductance (Table \ref{definitions}). The issue with PM is that
  $g_s$ is a function of carbon uptake, which is has a strong
  functional relation to ET through stomatal function. Therefore,
  when PM is formulated in terms of $g_s$, it is an implicit function
  of ET itself rather than an explicit function of ET. Here we will
  derive a new form of PM in which ET is an explicit function of plant
  parameters and environmental conditions, and use it to assess the
  ecosystem scale response to VPD (by taking a partial derivative).

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \footnotesize
  \begin{adjustbox}{center=15cm}
    \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{d \, e_s}{d \, T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citeA{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_{s}$  &  stomatal conductance & m s$^{-1}$
                                   & \citeA{Medlyn_2017} \\
    $g_{1}$  & ecosystem-scale slope parameter & Pa$^{0.5}$ & \citeA{Medlyn_2017} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda = \frac{\partial \; transpiration}{\partial\; CO_2\; assimilation}$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration &
                                                                      - & - \\
    GPP & gross primary production & $\mu$-mol C s$^{-1}$ m$^{-2}$ & -
    \\
    ET & evapotranspiration & W m$^{-2}$ & - \\
   uWUE & underlying water use efficiency & $\mu$-mol C Pa$^{0.5}$
                                            J$^{-1}$ ET  &
                                                           \citeA{Zhou_2015} \\
    \hline
  \end{tabular}
\end{adjustbox}
\end{table}

@@

\begin{table}
  \caption{Variable quantities in the ET response to VPD. Each value
    is varied to determine the effect of a range of expected plant and
    environmental conditions on ET response to VPD. A citation is
    provided for values of g$_1$ and uWUE, which are directly derived
    from previous literature. Conceptually, min. values are extracted
    from literature to correspond to approximately the 15th percentile
    of observed conditions during the growing season, med. values
    correspond to approximately the 50th percentile, and max. values
    correspond to approximately the 85th percentile. Values for $T$
    and g$_a$ are calculated from FLUXNET-2015 data (see supplemental
    material for description), rounding the 15th, 50th, and 85th
    percentile to the nearest 5$^o$C and 0.005 m/s, respectively.}
  \label{param_varying}
  \centering
  \footnotesize
  \begin{adjustbox}{center=15cm}
  \begin{tabular}{l c c c c c}
    \hline
    Symbol & units [units in citation] & min  & med & max
    & citation  \\
    \hline
g$_1$ & Pa$^{1/2}$ [kPa$^{1/2}$] & 63.25 [2.00] & 126.49 [4.00] & 189.74 [6.00] & Fig. 2, 7; \citeA{Medlyn_2017} \\
uWUE & $\mu$-mol C Pa$^{1/2}$ J$^{-1}$ ET [g C hPa$^{1/2}$ kg$^{-1}$ H$_2$O] & 2.33 [6.99] & 3.17 [9.52] & 4.01 [12.05] & Table 4; \citeA{Zhou_2015} \\
T & $^o$C & 10.00 & 20.00 & 30.00 & - \\
g$_a$ & m/s & 0.015 & 0.035 & 0.055 & - \\
    \hline
  \end{tabular}
  \end{adjustbox}
\end{table}

@@

\acknowledgments This work used eddy covariance data acquired and
shared by the FLUXNET community, including these networks: AmeriFlux,
AfriFlux, AsiaFlux, CarboAfrica, CarboEuropeIP, CarboItaly, CarboMont,
ChinaFlux, Fluxnet-Canada, GreenGrass, ICOS, KoFlux, LBA, NECC,
OzFlux-TERN, TCOS-Siberia, and USCCC. The ERA-Interim reanalysis data
are provided by ECMWF and processed by LSCE. The FLUXNET eddy
covariance data processing and harmonization was carried out by the
European Fluxes Database Cluster, AmeriFlux Management Project, and
Fluxdata project of FLUXNET, with the support of CDIAC and ICOS
Ecosystem Thematic Center, and the OzFlux, ChinaFlux and AsiaFlux
offices. We would like to thank Dr. Trevor Keenan for providing tools
for citation of the FLUXNET2015 dataset, available at
\url{https://github.com/trevorkeenan/FLUXNET_citations}. This material
is based upon work supported by the National Science Foundation
Graduate Research Fellowship under Grant No. DGE 16-44869 and the
National Aeronautics and Space Administration under grant
80NSSC18K0998. All code and data used in this analysis, including
those used to generate the figures and tables, are publicly available
at \url{https://github.com/massma/climate\_et}. We would like to thank
Emma Robinson and one anonymous reviewer for thoughtful comments that
greatly improved the strength of the manuscript.
